A Survey of Scientific Workflow Management Systems

### 2014-04-16 paper.tex
edits

\section{Introduction}

\vspace*{0.5cm}\hspace{1cm} Scientific Workflow Environments (SWE from here on)
are, in principle similar to the traditional programming environments
(programming language, interpreter or compiler, and runtime) albeit with a
larger scale of data and a higher abstraction of computation. The similarity is
such that scripting languages are sometimes advocated for programming at this
level. However, traditional scripts do not deal with parallelism and dataflow
aspects of workflow paradigm which makes them a strong contenders for the
purpose of high-level application programmers.
%
%Add more content here to beef up the context.
%
% Goals and benefits of a workflow management system
The goals of a workflow management system for scientific computing are as
follows:

@@

DAGs are by far, the most popular abstraction of the way workflows are
perceived and visualized. The MA
DAG\footnote{http://graal.ens-lyon.fr/diet/workflow.html}
\cite{caron-desprez:2005}, engine is one example of a DAG generator engine. MA
DAG has evolved from a regular DAG generator to a more abstract DAG generator
laced with control structures. MA DAG benefits from the workflow scheduling
mechanisms included in the middleware. Due to the impossibility to produce
complete DAGs owing to the presence of control structures such as conditionals
and loops, MA DAG generate partial DAGs greedily: a sub-DAG, as complete as
possible, is produced as soon as possible. Despite the presence of many
different approaches to workflow systems, new requirements
continue to pose workflow challenges. The community is yet to
realize a truly generic workflow language that can express a wide variety of
flow patterns encountered in scientific domains.

@@

The present work analyzes contemporary SWEs in order to find the capabilities
and key features in terms of the characteristics described in the introduction
part of this paper. The rest of this paper is organized as follows: Section
\ref{sec:overview} presents an overview of the key SWEs studied in this work.
Section \ref{sec:scal} discusses the optimization features and resulting
scalable performance as observed in the SWEs studied. Section \ref{sec:data}
studies the data management and description aspects of the SWEs. Section
\ref{sec:inter} presents a study of SWEs approaches to interface the
large-scale computing infrastructures. Section \ref{sec:lang} details the
language features and expressiveness of the SWEs studied. Section
\ref{sec:usable} discusses the usability aspects of the SWEs.  Section
\ref{sec:sum} presents a summary and a classification of all the aspects of
SWEs studied in this work. Throughout the paper, we also present tables
summarizing the aspects of SWEs considered in respective sections. It was not
possible to obtain an exhaustive information about all the SWEs considered in
this work since we rely on the peer reviewed publications for information on a
given SWE. Some features, such as extendability of SWE strongly depend on the
extension points designed in the code; consequently, we consider it present in
only those SWEs that have a documented description or demonstration of the
given feature.

\section{Overview of the Workflow Environments} \label{sec:overview}
In this section we present an overview of the Workflow Environments studied.
The paragraphs headers are organized by the name of the workflow environment
followed by the language they chiefly support separated by a forward slash (/).
The cases where the workflow language is not specifically defined or conversely
where there are multiple engines supporting a generic workflow language are
denoted by an asterisk (*) or a hyphen (-) at the respective places.
\paragraph{Taverna/scufl} The Taverna Workflow Management System
\cite{oinn-addis-etal:2004} is an open source scientific dataflow manager
developed by the mygrid~consortium in the UK.
%
Taverna includes a GUI-based rich-client workbench for workflow design and a
workflow enactment engine. Taverna can enact workflows expressed in the scufl
language. Scufl is a near Turing complete dataflow language with well-defined
semantics \cite{turi-missier-etal:2007,glatard-montagnat:2008}.

A Taverna workflow consists of a collection of processors connected by data
links, which establish a dependency between the output(s) of a processor and
the input(s) of another. Apart from data links, the processors may also be
connected through coordination links that make a processor invoke when a
previous processor ends its execution.  The processors are of different kinds
depending on the application code to be invoked: typically, processors are Web
Services, with input and output \textit{ports} that correspond to the
operation's invocations and response sequences defined in the service`s WSDL
interface.  Java classes and local scripts can be used as workflow processors,
as long as they expose their signature as input and output ports as required by
the model.  The engine orchestrates the execution of the processors in a way
that is consistent with the dependencies, and manages the flow of data through
the processors' ports. The overall workflow execution is data-driven, with a
push model: a processor's execution is started as soon as all of its inputs are
available.

@@

Taverna is not natively interfaced with a Grid infrastructure but the same can
be achieved with the help of extending the engine through one of its SPIs.

\paragraph{MOTEUR/scufl} MOTEUR \cite{glatard-montagnat-etal:2008} is a SWE
developed at the I3S laboratory, University of Nice at Sophia Antipolis (UNS).
MOTEUR facilitates executing workflows consisting of remote web- and
grid-services. It exploits service parallelism at workflow level and data
parallelism for multiple datasets.

MOTEUR has sophisticated data management mechanisms within the workflow
definition. It provides a diverse set of iterations strategies well adapted for
different combinations of input datasets at the processor. An iteration
strategy allows different combinations of the data arriving at a processor
along its multiple inputs. Running legacy application code on grids has been
eased by MOTEUR by providing a generic grid-service wrapper for these
applications \cite{glatard-montagnat-etal:2008a}. Grid-services thus created
can be readily deployed on the grids by the MOTEUR workflow enactor. MOTEUR
workflows are written in the scufl language and are compatible with those
written for the Taverna SWE. It also supports an extension of scufl language
called the Gwendia language, developed at the same laboratory. 
%
\paragraph{Makeflow} A make-like workflow platform~\cite{makeflow}.

%
\paragraph{Triana/xml} Triana \cite{taylor-shields-etal:2003} is a java-based
SWE developed at the Cardiff University. It consists of a workflow composition
GUI, a framework to integrate workflow with the different execution
environments and middlewares and a set of predefined services written as java
components. A Triana workflow is expressed as an XML grammar similar to a WSDL
specification for webservices \cite{taylor-wand-etal:2005}. A highlight of
Triana is that it provides ready-to-use interface for a wide variety of
middleware and distributed infrastructures including P2P, grids and
web-services frameworks. A Triana workflow provides dataflow as well as control
flow support. Triana allows conditional branching and looping in its workflows.

>>>>

\section{Introduction}
\label{sec:intro}
\vspace*{0.5cm}\hspace{1cm} Scientific Workflow Environments (SWE from here on)
are, in principle similar to the traditional programming environments
(programming language, interpreter or compiler, and runtime) albeit with a
larger scale of data and a higher abstraction of computation. The similarity is
such that scripting languages are sometimes advocated for programming at this
level. However, traditional scripts do not deal with parallelism and dataflow
aspects of workflow paradigm which makes them a strong contenders for the
purpose of high-level application programmers.
%
%Add more content here to beef up the context.
%
% Goals and benefits of a workflow management system
The goals of a workflow management system for scientific computing are as
follows:

@@

DAGs are by far, the most popular abstraction of the way workflows are
perceived and visualized. The MA
DAG~\cite{caron-desprez:2005},\km{could not find any recent paper on MA DAG, interesting concept} engine is one example of a DAG generator engine. MA
DAG has evolved from a regular DAG generator to a more abstract DAG generator
laced with control structures. MA DAG benefits from the workflow scheduling
mechanisms included in the middleware. Due to the impossibility to produce
complete DAGs owing to the presence of control structures such as conditionals
and loops, MA DAG generate partial DAGs greedily: a sub-DAG, as complete as
possible, is produced as soon as possible. Despite the presence of many
different approaches to workflow systems, new requirements
continue to pose workflow challenges. Some of the current challenges facing SWEs are:
\begin{enumerate}
    \item \textbf{Multisite execution.} Interfacing to existing computation and
        storate infrastructure is not trivial, not to mention the new and
        unforeseen access and compute modes. SWEs are yet to formulate a
        generalized interface towards the major platforms such that they will
        run on them out of the box.

    \item \textbf{Manysite execution.} Even if a SWE has interfaces built for
        more than one infrastructure, it is hard to optimally schedule large
        workflows to many remote sites simultaneously. The problem of limited
        control over remote sites and distributing interdependent data and
        tasks makes it non-trivial to efficiently load-balance work among them.

    \item \textbf{Expressivity.} Workflow languages have evolved over the years
        with reasonable expressivity being supported. However, they are often
        compared with those of the traditional programming languages. As of
        current state of the art, the workflow languages expressivity still
        have to catch up to the richness of programming languages across
        different generations such as C and python.  The community is yet to
        realize a truly generic workflow language that can express a wide
        variety of flow patterns encountered in scientific domains.

    \item \textbf{Interoperability.} Interoperable workflows will benefit to a
        broad audience of users. Many efforts at this are made in the recent
        past, such as by the consurtium projects and by individual players.
        However, a truly interoperable set of standards for workflows are yet
        to emerge.

@@

In the present work we survey and study the various contemporary SWEs in order
to find the capabilities and key features in terms of the characteristics
described in section~\ref{sec:intro}. We believe this study will serve to
address the following objectives:

\begin{enumerate}
  \item \textbf{Educate.} With more and more science being done with the help of
      computational methods, SWEs can add significant value to the process.
      They can accelerate the process, enable community driven science, drive
      the development at a manageable level of abstraction for users.
      However, scientific users are often not experienced programmers or
      compute savvy. This work will serve as a useful starting guide to
      educate science users of the available tools and techniques with
      features that might be suitable for their methods.

 \item \textbf{catalog.} We study a number of important features as offered by
     contemporary SWEs. This will help catalog and collect the SWEs by their
     offerings and identify mechanisms and needs for new and existing features.
     This will also help identify current trends in scientific computing
     community.

 \item \textbf{application.} We identify the applications and application
     classes well adapted to the workflow systems and are benefitting by them.
     This will help readers relate to the application they are associated to.
     We hope that readers can make a choice of which SWE to use if there is an
     adaptation scope.

 \item \textbf{evolution.} We comment on the evolution of the existing workflow
     systems and the possible current and future development directions which
     will help readers be aware of the general development direction of the
     SWEs.

The rest of this paper is organized as follows: Section
\ref{sec:overview} presents an overview of the key SWEs studied in this work.
Section \ref{sec:scal} discusses the optimization features and resulting
scalable performance as observed in the SWEs studied. Section \ref{sec:data}
studies the data management and description aspects of the SWEs. Section
\ref{sec:inter} presents a study of SWEs approaches to interface the
large-scale computing infrastructures. Section \ref{sec:lang} details the
language features and expressiveness of the SWEs studied. Section
\ref{sec:usable} discusses the usability aspects of the SWEs.  Section
\ref{sec:sum} presents a summary and a classification of all the aspects of
SWEs studied in this work. Throughout the paper, we also present tables
summarizing the aspects of SWEs considered in respective sections. It was not
possible to obtain an exhaustive information about all the SWEs considered in
this work. We rely on the following sources for our information: 
\begin{itemize}
    \item \textbf{Publications.} Peer reviewed publications in journals, conferences and workshops.
    \item \textbf{Technical Reports.} Technical reports and theses.
    \item \textbf{Web Presence.} Documentation and other information from the web pages of the host SWE.
    \item \textbf{Experience.} Where possible, we tried features of the SWEs by installing them on the available systems. 
\end{itemize}

Some features, such as extendability of SWE strongly depend on the extension
points designed in the code; consequently, we consider it present in only those
SWEs that have an explicit documented description or demonstration of the given
feature.

\section{Overview of the Workflow Environments} \label{sec:overview}
In this section we present an overview of the SWEs studied.  The paragraph
headers are organized by the name of the SWE followed by the name of internal
representation language use and support separated by a forward slash (/).  The
cases where an intermediate language is not specifically defined or conversely
where there are multiple systems supporting a generic representation are
denoted by an asterisk (*) or a hyphen (-) at the respective places.
\paragraph{Taverna/scufl} The Taverna Workflow Management System
\cite{oinn-addis-etal:2004} is an open source scientific dataflow manager
developed by the mygrid~consortium in the UK.
%
Taverna includes a GUI-based rich-client workbench for workflow design and a
workflow enactment runtime. Taverna can enact workflows expressed in the scufl
language. Scufl is a near Turing complete dataflow language with well-defined
semantics \cite{turi-missier-etal:2007,glatard-montagnat:2008}.

A Taverna workflow consists of a collection of processors connected by data
links, which establish a dependency between the output(s) of a processor and
the input(s) of another. Apart from data links, the processors may also be
connected through coordination links that make a processor invoke when a
previous processor ends its execution.  The processors are of different kinds
depending on the application code to be invoked: typically being services, with
input and output \textit{ports} that correspond to the operation's invocations
and response sequences defined in the service`s interface expressed as WSDL.
Java classes and local scripts can be used as workflow processors, as long as
they expose their signature as input and output ports as required by the model.
The engine orchestrates the execution of the processors in a way that is
consistent with the dependencies, and manages the flow of data through the
processors' ports. The overall workflow execution is data-driven, with a \emph{push}
model: a processor's execution is started as soon as all of its inputs are
available.

@@

Taverna is not natively interfaced with external computational infrastructure but the same can
be achieved with the help of extending the engine through one of its SPIs~\cite{maheshwari-missier-etal:2009}. 
\km{ToDo: Taverna current activity, applications support, GPU, Cloud support}

\paragraph{MOTEUR/scufl} MOTEUR \cite{glatard-montagnat-etal:2008} is a SWE
developed at the I3S laboratory, University of Nice at Sophia Antipolis (UNS).
MOTEUR facilitates executing workflows consisting of remote services. It
exploits service parallelism at workflow level and data parallelism for
multiple datasets.

MOTEUR has sophisticated language-level expressions for data management
mechanisms within a workflow definition. It provides a diverse set of
parallel iterators that are well adapted for different combinations of input datasets
at the processor. An iteration strategy allows different combinations of the
data arriving at a processor along its multiple inputs. Running legacy
application code on distributed systems has been eased by MOTEUR by providing a
generic grid-service wrapper for these applications
\cite{glatard-montagnat-etal:2008a}. Services thus created can be readily
deployed by the MOTEUR runtime. MOTEUR workflows are written in the scufl
language and are compatible with those written for the Taverna SWE. It also
supports an extension of scufl language called the Gwendia~\cite{gwendia}
language, developed at the same laboratory. 
%
\paragraph{Makeflow} A make-like workflow platform~\cite{makeflow}.

%
\paragraph{Triana/xml} Triana \cite{taylor-shields-etal:2003} is a java-based
SWE developed at the Cardiff University. It consists of a workflow composition
GUI, a framework to integrate workflow with the different execution
environments and middlewares and a set of predefined services written as java
components. A Triana workflow is expressed as an XML grammar similar to a WSDL
description for webservices \cite{taylor-wand-etal:2005}. A highlight of
Triana is that it provides ready-to-use interface for a wide variety of
middleware and distributed infrastructures including P2P, grids and
web-services frameworks. A Triana workflow provides dataflow as well as control
flow support. Triana allows conditional branching and looping in its workflows.
